This chapter provides the relevant background information. It largely



We explore two lineages of resourceful type systems. In the first, modern
resourceful type systems trace their roots to Girard's Linear Logic~\cite{}
which was one of the first treatments of data as a resource inside a program. Bounded Linear Logic (BLL)
developed this idea further, refining the coarse grained view of data as either
linear or non linear. Several subsequent works generalised BLL, coalescing into
the notion of a \textit{graded} type system in languages such as
Granule~\cite{}. This lineage treats these systems essentially as refinements of an
underlying linear structure.

At the same time that these systems were being studied, resourceful types were
also being approached from an entirely different perspective --- that of
computational effects. \jnote{more to go here}.

\begin{table}[htbp]
\centering
\caption{Timeline.}
\label{tbl:1}
\begin{tabular}{r l l}
  Year & \ \ Linear & \ \ Graded \\
  \toprule
  1986 & & \textasteriskcentered\ D.K. Gifford, J.M. Lucassen \\
       & & \ \ \textit{Integrating Functional and} \\
       & & \ \ \textit{Imperative Programming}  \\

  1987 & \textasteriskcentered\ J.Y. Girard \\
       & \ \ \textit{Linear Logic} & \\

  1990 & & \textasteriskcentered\ E. Moggi \\
       & & \ \ \textit{Notions of Computation and Monads} \\

  1991 & \textasteriskcentered\ J.Y. Girard et al. & \\
       & \ \ \textit{Bounded Linear Logic} & \\

  1994 & \textasteriskcentered\ J.S. Hodas, D. Miller &  \\
       & \ \ \textit{Logic Programming in a Fragment} & \\
       & \ \ \textit{of Intuitionistic Linear Logic} & \\

  2000 & & \textasteriskcentered\ P. Wadler, P. Thiemann \\
       & & \ \ \textit{Marriage of effects and monads} \\
  2011 & \textasteriskcentered\ U.D. Lago, M. Gaboardi &  \\
       & \ \ \textit{Linear Dependent Types} & \\
       & \ \ \textit{and Relative Completeness} & \\
  2013 & & \textasteriskcentered\ T. Petricek et al. \\
       & & \ \ \textit{Coeffects: Unified Static} \\
       & & \ \ \textit{Analysis of Context-Dependence} \\
  2014 & \textasteriskcentered\ D.R. Ghica, A. I. Smith & \textasteriskcentered\
  T. Petricek et al. \\
       & \ \ \textit{Bounded Linear Types in} & \ \ \textit{Coeffects: a
                                                Calculus of} \\
       & \ \ \textit{a Resource Semiring}  & \ \ \textit{Context-Dependent Computation} \\
       & \textasteriskcentered\ A. Brunel et al. & \\
       & \ \ \textit{A Core Quantative Coeffect Calculus} & \\
  2016 & \textasteriskcentered\ M. Gaboardi et al. & \textasteriskcentered\ C. McBride \\
       & \ \ \textit{Combining Effects and} & \ \ \textit{I Got Plenty o' Nuttin'} \\
       & \ \ \textit{Coeffects via Grading} & \\
  2017 & & \textasteriskcentered\ J.P. Bernardy et al. \\
       & & \ \ \textit{Linear Haskell} \\
  2018 & & \textasteriskcentered\ R. Atkey \\
       & & \ \ \textit{Syntax and Semantics of} \\
       & & \ \ \textit{Quantative Type Theory} \\
  2019 & \textasteriskcentered\ Orchard et al. & \\
       & \ \ \textit{Quantitative Program Reasoning} & \\
       & \ \ \textit{with Graded Modal Types} & \\
  % 2020 & LOPSTR &  \\
  % 2021 & & \textasteriskcentered\ A. Abel, J.P. Bernardy \\
  %      & & \ \ \textit{A Unified View of Modalities} \\
  %      & & \ \ \textit{in Type Systems} \\
  %      & & \textasteriskcentered\ P. Choudhury et al. \\
  %      & & \ \ \textit{A Graded Dependent Type System} \\
  %      & & \ \ \textit{with a Usage-Aware Semantics} \\
  %      & & \textasteriskcentered\ B. Moon et al. \\
  %      & & \ \ \textit{Graded Modal Dependent} \\
  %      & & \ \ \textit{Type Theory} \\
  %      & & \textasteriskcentered\ E. Brady \\
  %      & & \ \ \textit{Idris 2: Quantative Type} \\
  %      & & \ \ \textit{Theory in Practice} \\
  %
  \bottomrule
\end{tabular}
\end{table}


\paragraph{Terminology}
Before delving into linear and graded systems, we briefly frame the approach
we will take to discussing the relevant background material. Throughout this
chapter (and subsequent chapters) we will tend towards using a
\textit{types-and-programs} terminology rather than
\textit{propositions-and-proofs}. Through the lens of the Curry-Howard
correspondence, one can switch smoothly to viewing our approach to program
synthesis as proof search in logic.

The functional programming languages we discuss are presented as typed
calculi given by sets of \textit{types}, \textit{terms} (programs), and \textit{typing
 rules} that relate a term to its type. The most well-known typed
calculus is the simply-typed $\lambda$-calculus, which corresponds to
intuitionistic logic.

A \textit{judgment} defines the typing relation between a type and a term based on a
\textit{context}. In the simple typed $\lambda$-calculus, judgments have the
form: $\Gamma \vdash t : A$, stating that under some context of
\textit{assumptions} $\Gamma$ the program term $t$ can be assigned the type $A$.
An assumption is a name with an associated type, written $x : A$ and
corresponds to an in-scope variable in a program.

A term can be related to a type if we can derive a valid judgment through the
application of typing rules. The application of these rules forms a tree
structure known as a \textit{typing derivation}.


\section{Linear and substructural logics}
Linear logic~\cite{} was introduces by Girard as a way of being more descriptive
about the properties of a derivation in intuitionistic logic. In type systems
such as the simply typed $\lambda$-calculus, the properties of \textit{weakening},
\textit{contraction}, and \textit{exchange} are assumed implicitly. These are
typing rules which are \textit{structural} as they determine how the context
may be used rather than being directed by the syntax. Weakening is a rule which allows terms
that are not needed in a typing derivation to be discarded. Contraction works as a dual to weakening, allowing an
assumption in the context to be used more than once. Finally, exchange allows assumptions in a context to arbitrarily re-ordered.


\begin{figure}[H]
  \begin{align*}
  \begin{array}{c}
    \inferrule*[right=Weakening]{\Gamma \vdash t : B}{\Gamma, x : A \vdash t : B}
    \;\;\;\;
    \;\;\;\;
    \;\;\;\;
    \inferrule*[right=Contraction]{\Gamma, x : A, y : A \vdash t : B}{\Gamma, x : A \vdash t : B}
    \\\\
    \inferrule*[right=Exchange]{\Gamma_{1}, y : B, x : A, \Gamma_{2} \vdash t : C} {\Gamma_{1}, x : A, y : B, \Gamma_{2} \vdash t : C}
    \end{array}
  \end{align*}
\end{figure}

Linear logic is known as a \textit{substructural} logic because it lacks the weakening
and contraction rules, while permitting exchange.

The disallowance of these rules means that in order to construct of a typing
derivation, each assumption must be used exactly once --- arbitrarily copying or
discarding values is disallowed, excluding a vast number of programs from being
typeable in linear logic. Non-restricted usage of a value is recovered through
the modal operator $!$ (also called ``bang'', ``of-course'', or the
\textit{exponential} modality). Affixing $!$ to a type captures the notion that values of
that type may be used freely in a program.

providing a binary view of data as a
resource inside a program: values are either linear or completely unrestricted.

Bounded Linear Logic, took this idea further --- instead of a single modal
operator, $!$ is replaced with a family of modal operators indexed by terms
which provide an upper bound on usage~\cite{}. These terms provide an upper bound on
the usage of a values inside term, e.g. $!_{3}A$ is the type of $A$ values which
may be used up to 3 times.


\jnote{more about Hodas and Miller, ICFP and Granule etc}
The idea of data as a resource inside a program has been extended further by the
proliferation of graded type systems.


\subsection{A graded linear typing calculus}
\label{sec:linear-base}

Having seen the resourceful types from linear logic to graded type systems, we
are now in a position to define a full typing calculus, based on the linear
$\lambda$-calculus extended with a graded modal type. This calculus is
equivalent to the core calculus Granule,
\textsc{GrMini}\cite{}. Granule's full type system is an extension of this
graded linear core, with the addition of polymorphism, indexed-types, pattern
matching, and the ability to use multiple different graded modalities. We refer
to this system as the \textit{linear-base} calculus, reflecting the underlying
linear structure of the system.

The types of the linear-base calculus are defined as:
\begin{align*}
\hspace{-0.9em}\GRANULEnt{A} , \GRANULEnt{B} & ::=
       \GRANULEnt{A} \multimap \GRANULEnt{B}
  \mid \Box_{  \GRANULEnt{r}  }  \GRANULEnt{A}
{\small{\tag{types}}}
\end{align*}
The type $\Box_{r} A$ is an indexed family of type operators where $r$ is a
\textit{grade} ranging over the elements of a pre-ordered semiring
$({\mathcal{R}}, {*}, {1}, {+}, {0},
{\sqsubseteq})$ parameterising the calculus (where $\ast$ and $+$
are monotonic with respect to the pre-order $\sqsubseteq$).

The syntax of terms provides the elimination and introduction forms:
\begin{align*}
\hspace{-0.8em} \GRANULEnt{t} ::= \;
       & \GRANULEmv{x}
  \mid \lambda  \GRANULEmv{x}  .  \GRANULEnt{t}
  \mid \GRANULEnt{t_{{\mathrm{1}}}} \, \GRANULEnt{t_{{\mathrm{2}}}}
  \mid \GRANULEsym{[}  \GRANULEnt{t}  \GRANULEsym{]}
  \mid \textbf{let} \, [  \GRANULEmv{x}  ] =  \GRANULEnt{t_{{\mathrm{1}}}}  \, \textbf{in} \,  \GRANULEnt{t_{{\mathrm{2}}}}
{\small{\tag{terms}}}
\end{align*}
In addition the the terms of the linear $\lambda$-calculus, we also have the
construct $\GRANULEsym{[}  \GRANULEnt{t}  \GRANULEsym{]}$ which introduces a graded modal type $\Box_{  \GRANULEnt{r}  }  \GRANULEnt{A}$ by `promoting' a term $t$ to the graded modality, and it's dual $\textbf{let} \, [  \GRANULEmv{x}  ] =  \GRANULEnt{t_{{\mathrm{1}}}}  \, \textbf{in} \,  \GRANULEnt{t_{{\mathrm{2}}}}$ eliminates a graded modal value $\GRANULEnt{t_{{\mathrm{1}}}}$, binding a graded variable $x$
in scope of $\GRANULEnt{t_{{\mathrm{2}}}}$. The typing rules provide further understanding of the
behaviour of these terms.

Typing judgments are of the form $\Gamma  \vdash  \GRANULEnt{t}  :  \GRANULEnt{A}$, where $\Gamma$ ranges over contexts:
\begin{equation*}
  \Gamma ::= \emptyset
  \mid \Gamma  \GRANULEsym{,}   \GRANULEmv{x}  :  \GRANULEnt{A}
  \mid \Gamma  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{r}  } }   \GRANULEnt{A}
\tag{contexts}
\end{equation*}

Thus, a context may be empty $\emptyset$, extended with a
linear assumption $\GRANULEmv{x}  :  \GRANULEnt{A}$ or extended with a graded assumption $\GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{r}  } }   \GRANULEnt{A}$. For linear assumptions, structural rules of weakening
and contraction are disallowed. Graded assumptions may be used
non-linearly according to the constraints given by their grade, the
semiring element $r$. Throughout, comma denotes disjoint context
concatenation.

Various operations on contexts are used to capture non-linear data flow
via grading. Firstly, \emph{context addition}~\eqref{def:contextAdd} provides an
analogue to contraction, combining contexts that have come from
typing multiple subterms in a rule.
Context addition, written $\Gamma_{{\mathrm{1}}}  \GRANULEsym{+}  \Gamma_{{\mathrm{2}}}$, is undefined if $\Gamma_{{\mathrm{1}}}$ and $\Gamma_{{\mathrm{2}}}$
overlap in their linear assumptions. Otherwise graded assumptions appearing
in both contexts are combined via the semiring $+$ of their grades.

\begin{definition}[Context addition]\label{def:contextAdd}

\begin{align*}
    \setlength{\arraycolsep}{0.1em}
    \begin{array}{cc}
    \begin{array}{rl}
    \GRANULEsym{(}  \Gamma  \GRANULEsym{,}   \GRANULEmv{x}  :  \GRANULEnt{A}   \GRANULEsym{)}  \GRANULEsym{+}  \Gamma' & = \GRANULEsym{(}  \Gamma  \GRANULEsym{+}  \Gamma'  \GRANULEsym{)}  \GRANULEsym{,}   \GRANULEmv{x}  :  \GRANULEnt{A} \quad \text{iff} \,\; x
    \not\in | \Gamma' | \\
    \Gamma  \GRANULEsym{+}  \GRANULEsym{(}  \Gamma'  \GRANULEsym{,}   \GRANULEmv{x}  :  \GRANULEnt{A}   \GRANULEsym{)} & = \GRANULEsym{(}  \Gamma  \GRANULEsym{+}  \Gamma'  \GRANULEsym{)}  \GRANULEsym{,}   \GRANULEmv{x}  :  \GRANULEnt{A} \quad \text{iff} \,\; x \not\in | \Gamma | \\
    \GRANULEsym{(}  \Gamma  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{r}  } }   \GRANULEnt{A}   \GRANULEsym{)}  \GRANULEsym{+}  \GRANULEsym{(}  \Gamma'  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{s}  } }   \GRANULEnt{A}   \GRANULEsym{)} & = \GRANULEsym{(}  \Gamma  \GRANULEsym{+}  \Gamma'  \GRANULEsym{)}  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEsym{(}  \GRANULEnt{r}  \GRANULEsym{+}  \GRANULEnt{s}  \GRANULEsym{)}  } }   \GRANULEnt{A}
    \end{array}
      \;\;\quad & \quad\;\;
    \begin{array}{rl}
      \emptyset   \GRANULEsym{+}  \Gamma & = \Gamma \\
      \Gamma  \GRANULEsym{+}   \emptyset & = \Gamma \\ \\
    \end{array}
  \end{array}
  \end{align*}

Note that this is a declarative specification of context addition. Graded
assumptions may appear in any position in $\Gamma$ and $\Gamma'$ as witnessed by
the algorithmic specification where for all $\Gamma_{{\mathrm{1}}}, \Gamma_{{\mathrm{2}}}$
  \emph{context addition} is defined
as follows by ordered cases matching inductively on the structure of
$\Gamma_{{\mathrm{2}}}$:
\begin{align*}
\Gamma_{{\mathrm{1}}}  \GRANULEsym{+}  \Gamma_{{\mathrm{2}}} = \left\{\begin{matrix}
    \begin{array}{ll}
    \Gamma_{{\mathrm{1}}} &
     \Gamma_{{\mathrm{2}}} = \emptyset
             \\
      ((\Gamma'_{{\mathrm{1}}}, \Gamma''_{{\mathrm{1}}}) + \Gamma'_{{\mathrm{2}}}), \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEsym{(}  \GRANULEnt{r}  \GRANULEsym{+}  \GRANULEnt{s}  \GRANULEsym{)}  } }   \GRANULEnt{A} \; &
\Gamma_{{\mathrm{2}}} = \Gamma'_{{\mathrm{2}}}  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{s}  } }   \GRANULEnt{A} \wedge \Gamma_{{\mathrm{1}}} = \Gamma'_{{\mathrm{1}}}  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{r}  } }   \GRANULEnt{A},\Gamma''_{{\mathrm{1}}}  \\
 (\Gamma_{{\mathrm{1}}} + \Gamma'_{{\mathrm{2}}}), \GRANULEmv{x}  :  \GRANULEnt{A} & \Gamma_{{\mathrm{2}}} = \Gamma'_{{\mathrm{2}}}  \GRANULEsym{,}   \GRANULEmv{x}  :  \GRANULEnt{A}\ \wedge\  \GRANULEmv{x}  :  \GRANULEnt{A} \notin \Gamma_{{\mathrm{1}}}
    \end{array}
  \end{matrix}\right.
\end{align*}
\end{definition}

\begin{figure}[H]
\hspace{-0.5em}
\begin{align*}
\hspace{-0.5em}
  \begin{array}{c}
  \inferrule*[right = Var]
  {\;}
  {\GRANULEmv{x}  :  \GRANULEnt{A}   \vdash  \GRANULEmv{x}  :  \GRANULEnt{A}}
\;\;
  \inferrule*[right = Abs]
  {\Gamma  \GRANULEsym{,}   \GRANULEmv{x}  :  \GRANULEnt{A}   \vdash  \GRANULEnt{t}  :  \GRANULEnt{B}}
  {\Gamma  \vdash   \lambda  \GRANULEmv{x}  .  \GRANULEnt{t}   :  \GRANULEnt{A}  \rightarrow  \GRANULEnt{B}}
\;\;
  \inferrule*[right = App]
  {\Gamma_{{\mathrm{1}}}  \vdash  \GRANULEnt{t_{{\mathrm{1}}}}  :  \GRANULEnt{A}  \rightarrow  \GRANULEnt{B} \;\;\;
   \Gamma_{{\mathrm{2}}}  \vdash  \GRANULEnt{t_{{\mathrm{2}}}}  :  \GRANULEnt{A} }
  {\Gamma_{{\mathrm{1}}}  \GRANULEsym{+}  \Gamma_{{\mathrm{2}}}  \vdash  \GRANULEnt{t_{{\mathrm{1}}}} \, \GRANULEnt{t_{{\mathrm{2}}}}  :  \GRANULEnt{B}}
\\[0.75em]
 \inferrule*[right = Weak]
  {\Gamma  \vdash  \GRANULEnt{t}  :  \GRANULEnt{A}}
  {\Gamma  \GRANULEsym{,}    [  \Delta  ]_{  0  }    \vdash  \GRANULEnt{t}  :  \GRANULEnt{A}}
\;\;\;
\inferrule*[right = Der]
  {\Gamma  \GRANULEsym{,}   \GRANULEmv{x}  :  \GRANULEnt{A}   \vdash  \GRANULEnt{t}  :  \GRANULEnt{B}}
  {\Gamma  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{   1   } }   \GRANULEnt{A}   \vdash  \GRANULEnt{t}  :  \GRANULEnt{B}}
\;\;\;
\inferrule*[right = Approx]
{\Gamma  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{r}  } }   \GRANULEnt{A}    \GRANULEsym{,}  \Gamma'  \vdash  \GRANULEnt{t}  :  \GRANULEnt{A} \quad r \sqsubseteq s }
{\Gamma  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{s}  } }   \GRANULEnt{A}    \GRANULEsym{,}  \Gamma'  \vdash  \GRANULEnt{t}  :  \GRANULEnt{A}}
\\[0.75em]
\inferrule*[right = Pr]
  {\GRANULEsym{[}  \Gamma  \GRANULEsym{]}  \vdash  \GRANULEnt{t}  :  \GRANULEnt{A}}
  {\textcolor{coeffectColor}{ \GRANULEnt{r}   \textcolor{coeffectColor}{\,\cdot\,} }   \GRANULEsym{[}  \Gamma  \GRANULEsym{]}    \vdash  \GRANULEsym{[}  \GRANULEnt{t}  \GRANULEsym{]}  :   \Box_{  \GRANULEnt{r}  }  \GRANULEnt{A}}
\;\;\;
\inferrule*[right = Let$\Box$]
  {\Gamma_{{\mathrm{1}}}  \vdash  \GRANULEnt{t_{{\mathrm{1}}}}  :   \Box_{  \GRANULEnt{r}  }  \GRANULEnt{A} \;\;\;
   \Gamma_{{\mathrm{2}}}  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{r}  } }   \GRANULEnt{A}   \vdash  \GRANULEnt{t_{{\mathrm{2}}}}  :  \GRANULEnt{B} }
    {\Gamma_{{\mathrm{1}}}  \GRANULEsym{+}  \Gamma_{{\mathrm{2}}}  \vdash   \textbf{let} \, [  \GRANULEmv{x}  ] =  \GRANULEnt{t_{{\mathrm{1}}}}  \, \textbf{in} \,  \GRANULEnt{t_{{\mathrm{2}}}}   :  \GRANULEnt{B}}
\end{array}
\end{align*}
\vspace{-1.25em}
  \caption{Typing rules of the graded linear $\lambda$-calculus}
\label{fig:typing}
\vspace{-0.65em}
 \end{figure}


Figure~\ref{fig:typing} defines the typing rules.
Linear variables are typed in a singleton context
(\textsc{Var}). Abstraction (\textsc{Abs}) and application (\textsc{App})
follow the rules of the linear $\lambda$-calculus.
The $\textsc{Weak}$ rule captures
weakening of assumptions graded by $0$ (where $[  \Delta  ]_{  0  }$ denotes a context
containing only graded assumptions graded by $0$). Context addition and
\textsc{Weak} together therefore provide the rules of substructural rules of contraction
and weakening.
Dereliction ($\textsc{Der}$),
allows a linear assumption to be converted to a graded assumption with grade
$1$. Grade approximation is captured by the $\textsc{Approx}$
rule, which allows a grade $s$ to be converted to another grade $r$,
providing that $r$ \textit{approximates} $s$, where the relation
$\sqsubseteq$ is the pre-order provided
with the semiring.
Introduction and elimination of the graded modality is provided by the
$\textsc{Pr}$ and $\textsc{Let}$ rules
respectively. The $\textsc{Pr}$ rule propagates the grade $r$ to the
assumptions through \emph{scalar multiplication} of $\Gamma$ by $r$ where
every assumption in $\Gamma$ must already be graded (written $\GRANULEsym{[}  \Gamma  \GRANULEsym{]}$ in the rule), given by definition~\eqref{def:scalar}.
%
%
\begin{definition}[Scalar context multiplication]
  \label{def:scalar}
 A context which consists solely of graded assumptions can be multiplied by a
 semiring grade $r \in \mathcal{R}$
\begin{align*}
   \textcolor{coeffectColor}{ \GRANULEnt{r}   \textcolor{coeffectColor}{\,\cdot\,} }   \emptyset = \emptyset
    \qquad\qquad
    \textcolor{coeffectColor}{ \GRANULEnt{r}   \textcolor{coeffectColor}{\,\cdot\,} }  \GRANULEsym{(}  \Gamma  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{s}  } }   \GRANULEnt{A}   \GRANULEsym{)} = \GRANULEsym{(}   \textcolor{coeffectColor}{ \GRANULEnt{r}   \textcolor{coeffectColor}{\,\cdot\,} }  \Gamma   \GRANULEsym{)}  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEsym{(}  \GRANULEnt{r}  \textcolor{coeffectColor}{\,\cdot\,}  \GRANULEnt{s}  \GRANULEsym{)}  } }   \GRANULEnt{A}
\end{align*}
\end{definition}

The $\textsc{Let}$ rule eliminates a graded modal value $\Box_{  \GRANULEnt{r}  }  \GRANULEnt{A}$
into a graded assumption $\GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{r}  } }   \GRANULEnt{A}$ with a matching
grade in the scope of the \textbf{let} body. This is also referred to as
``unboxing''.


We give an example of graded modalities using a graded modality indexed
by the semiring of natural numbers.

%SKI
\begin{example}
\label{ex:s-comb}
  The natural number semiring with discrete ordering
  $(\mathbb{N}, \ast, 1, +, 0, \equiv)$ provides a graded modality
  that counts exactly how many times non-linear values are used. As a
  simple example, the \emph{S} combinator from the SKI system of combinatory logic is typed and defined:
% then2 : forall {a b c : Type} . (a -> (b -> c)) -> (a -> b) -> (a
% [2] -> c)
\begin{align*}
s & : \GRANULEsym{(}  \GRANULEnt{A}  \rightarrow  \GRANULEsym{(}  \GRANULEnt{B}  \rightarrow  \GRANULEnt{C}  \GRANULEsym{)}  \GRANULEsym{)}  \rightarrow   \GRANULEsym{(}  \GRANULEnt{A}  \rightarrow  \GRANULEnt{B}  \GRANULEsym{)}  \rightarrow  \GRANULEsym{(}    \Box_{  \GRANULEsym{2}  }  \GRANULEnt{A}    \rightarrow  \GRANULEnt{C}  \GRANULEsym{)} \\
s & = \lambda  \GRANULEmv{x}  .    \lambda  \GRANULEmv{y}  .    \lambda  \GRANULEmv{z'}  .    \textbf{let} \, [  \GRANULEmv{z}  ] =  \GRANULEmv{z'}  \, \textbf{in} \,   \GRANULEsym{(}  \GRANULEmv{x} \, \GRANULEmv{z}  \GRANULEsym{)} \, \GRANULEsym{(}  \GRANULEmv{y} \, \GRANULEmv{z}  \GRANULEsym{)}
\end{align*}
The graded modal value $z'$ captures the `capability' for a value
of type $A$ to be used twice. This capability is made available by eliminating
$\Box$ (via \textbf{let}) to the variable $z$, which is
graded $z : [A]_2$ in the scope of the body.
\end{example}



\paragraph{Metatheory}
The admissibility of substitution is a key result that holds
for this language~\cite{DBLP:journals/pacmpl/OrchardLE19}, which is
leveraged in soundness of the synthesis calculi.
%
\begin{restatable}[Admissibility of substitution]{lemma}{linearSubst}
Let $\Delta  \vdash  \GRANULEnt{t'}  :  \GRANULEnt{A}$, then:
\label{lemma:substitution}
\begin{itemize}[leftmargin=1em]
\item (Linear) \hspace{0.04em} If $\Gamma  \GRANULEsym{,}   \GRANULEmv{x}  :  \GRANULEnt{A}    \GRANULEsym{,}  \,  \GRANULEsym{,}   \Gamma'   \vdash  \GRANULEnt{t}  :  \GRANULEnt{B}$ then $\Gamma  \GRANULEsym{+}  \Delta  \GRANULEsym{+}  \Gamma'  \vdash   [  \GRANULEnt{t'}  /  \GRANULEmv{x}  ]  \GRANULEnt{t}   :  \GRANULEnt{B}$
\item (Graded) If $\Gamma  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{r}  } }   \GRANULEnt{A}    \GRANULEsym{,}  \,  \GRANULEsym{,}   \Gamma'   \vdash  \GRANULEnt{t}  :  \GRANULEnt{B}$
then $\Gamma  \GRANULEsym{+}  \GRANULEsym{(}   \textcolor{coeffectColor}{ \GRANULEnt{r}   \textcolor{coeffectColor}{\,\cdot\,} }  \Delta   \GRANULEsym{)}  \GRANULEsym{+}  \Gamma'  \vdash   [  \GRANULEnt{t'}  /  \GRANULEmv{x}  ]  \GRANULEnt{t}   :  \GRANULEnt{B}$
\end{itemize}
\end{restatable}

\section{Graded types: An alternative history}
\jnote{talk about lineage of graded type systems}

\subsection{Graded-base}
\label{sec:graded-base}

We now define a core calculus for a fully graded type system. This system is much closer
to those outlined above than the linear-base calculus, drawing from the coeffect calculus of \citet{petricek2014coeffects}, Quantitative Type Theory (QTT) by \citet{McBride2016} and refined further by \citet{quantitative-type-theory} (although we omit dependent types from our language), the calculus of \citet{DBLP:journals/pacmpl/AbelB20}, and other graded dependent type theories~\cite{quantitative-type-theory,DBLP:conf/esop/MoonEO21}. Similar
systems also form the basis of the core of the linear types extension to
Haskell~\cite{DBLP:journals/pacmpl/BernardyBNJS18}. We refer to this system as
the \textit{graded-base} calculus to differentiate it from linear-base.

The syntax of graded-base types is given by:
\begin{align*}
\hspace{-0.9em}\GRANULEnt{A} , \GRANULEnt{B} & ::=
       \GRANULEnt{A} ^ \GRANULEnt{r}  \rightarrow  \GRANULEnt{B}
  \mid \Box_{  \GRANULEnt{r}  }  \GRANULEnt{A}
{\small{\tag{\textit{types}}}}
\end{align*}
where the function space $\GRANULEnt{A} ^ \GRANULEnt{r}  \rightarrow  \GRANULEnt{B}$ annotates the input type with a \emph{grade} $\GRANULEnt{r}$
drawn from a pre-ordered semiring
$(\mathcal{R}, {\ast}, {1}, {+}, {0}, \sqsubseteq)$ which paramterising the
calculus as in linear-base.

The graded necessity modality $\Box_{  \GRANULEnt{r}  }  \GRANULEnt{A}$ is similarly annotated by the grade
$\GRANULEnt{r}$ being an element of the semiring.


The syntax of terms is given as:
%
\begin{align*}
\hspace{-0.8em} \GRANULEnt{t} ::= \;
       & \GRANULEmv{x}
  \mid \lambda  \GRANULEmv{x}  .  \GRANULEnt{t}
  \mid \GRANULEnt{t_{{\mathrm{1}}}} \, \GRANULEnt{t_{{\mathrm{2}}}}
  \mid \GRANULEsym{[}  \GRANULEnt{t}  \GRANULEsym{]}
  % \mid \textbf{case} \  \GRANULEnt{t}  \ \textbf{of} \   \GRANULEnt{p_{{\mathrm{1}}}}  \mapsto  \GRANULEnt{t_{{\mathrm{1}}}} ; ... ;  \GRANULEnt{p_{\GRANULEmv{n}}}  \mapsto  \GRANULEnt{t_{\GRANULEmv{n}}}
{\small{\tag{\textit{terms} }}}
\end{align*}
%
Similarly to linear-base, terms consist of a graded $\lambda$-calculus, extended
with a \textit{promotion} construct [t] which introduces a graded modality
explicitly.


\begin{definition}[Context addition]\label{def:contextAdd}

\begin{align*}
\Gamma_{{\mathrm{1}}}  \GRANULEsym{+}  \Gamma_{{\mathrm{2}}} = \left\{\begin{matrix}
    \begin{array}{ll}
    \Gamma_{{\mathrm{1}}} &
    \Gamma_{{\mathrm{2}}} = \emptyset
             \\
      ((\Gamma'_{{\mathrm{1}}}, \Gamma''_{{\mathrm{1}}}) + \Gamma'_{{\mathrm{2}}}), \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEsym{(}  \GRANULEnt{r}  \GRANULEsym{+}  \GRANULEnt{s}  \GRANULEsym{)}  } }   \GRANULEnt{A} \; &
\Gamma_{{\mathrm{2}}} = \Gamma'_{{\mathrm{2}}}  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{s}  } }   \GRANULEnt{A} \wedge \Gamma_{{\mathrm{1}}} = \Gamma'_{{\mathrm{1}}}  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{r}  } }   \GRANULEnt{A},\Gamma''_{{\mathrm{1}}} \\
 \GRANULEsym{(}  \Gamma_{{\mathrm{1}}}  \GRANULEsym{+}  \Gamma'_{{\mathrm{2}}}  \GRANULEsym{)}  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{s}  } }   \GRANULEnt{A} & \Gamma_{{\mathrm{2}}} = \Gamma'_{{\mathrm{2}}}  \GRANULEsym{,}   \GRANULEmv{x}  :_{\textcolor{coeffectColor}{  \GRANULEnt{s}  } }   \GRANULEnt{A} \wedge \GRANULEmv{x} \not\in \mathsf{dom}(\Gamma_{{\mathrm{1}}})
\end{array}
  \end{matrix}\right.
\end{align*}
\end{definition}

\begin{figure}[H]
\hspace{-0.5em}
\begin{align*}
\hspace{-0.5em}
\begin{array}{c}
\GRANULEdruleTyVar{}
\;\;\;
\GRANULEdruleTyAbs{}
\;\;\;
\GRANULEdruleTyApp{}
\\\\
\GRANULEdruleTyPr{}
\;\;\;
\GRANULEdruleTyApprox{}
\end{array}
\end{align*}
\vspace{-0.5em}
\caption{Typing rules for graded-base}
\label{fig:typing}
\vspace{-0.5em}
 \end{figure}

Figure~\ref{fig:typing} gives the full typing rules, which helps explain the meaning of
the syntax with reference to their static semantics.

Variables (rule \textsc{Var}) are
typed in a context where the variable $x$ has grade $1$ denoting its
single usage here. All other variable assumptions are given the grade
of the $0$ semiring element (providing \emph{weakening}),
using \textit{scalar multiplication} of contexts by a grade, re-using definition~\eqref{def:scalar}.


\section{Two typed calculi}


Having outlined the two major lineages of resourceful types, we
are now left with the question: what approach should we use as the basis of our
program synthesis tool? Both approaches pose there own challenges and questions
which influence the design of a synthesis calculus. The breadth of these
differences leaves omitting either approach an undesirable prospect. For this
reason, this thesis tackles languages based on both approaches:

\begin{enumerate}
        \item We begin with a simplified synthesis calculus for a core language
        based on the linear $\lambda$ calculus outlined in
        section~\eqref{sec:linear-base}. To better illustrate the practicality
        of the syntesis calculus, the language is extended with product
        and sum types, as well as a unit type.
        \item
\end{enumerate}



% The different approaches to graded types here gives rise to the question: how might the
% features of the target language influence the design of a synthesis tool?

This chapter introduced some of the key features of languages with resourceful
types. Linear and graded types embed usage constraints in the typing rules,
enforcing the notion that a well-typed program is also \textit{well-resourced}.

The next chapter focuses on the linear-base core calculus of
section~\ref{sec:linear-base}, extending this calculus with multiplicative and additive
types, as well as a unit type to form a more practical programming language. This
then comprises the target language of a synthesis algorithm. Likewise, the graded-base
calculus is revisited in chapter~\ref{chapter:graded-base}, where it is extended
with (G)ADTs, and recursion, providing a target language for a more in-depth and
featureful synthesis tool.
